\documentclass[12px]{report}

\usepackage{xcolor}
\usepackage[style=numeric,sorting=none,backend=biber]{biblatex}
\addbibresource{references.bib} % The file containing our references, in BibTeX format
\usepackage{csquotes} % Recommended by biblatex
\usepackage[colorlinks=true,linkcolor=violet]{hyperref}
\usepackage{booktabs}
\usepackage{float}
\usepackage[all]{hypcap} %jump to img instead of caption text

\usepackage{listings}
\lstset{
    language=xml,
    basicstyle=\footnotesize,
    stringstyle={\ttfamily \color{teal}},
    keywordstyle={\bfseries \color{blue}},
    showstringspaces=false,
    frame=tb,
    morekeywords={
        Bundle, 
        Composition, 
        Patient, 
        Organization, 
        Condition, 
        Questionnaire, 
        QuestionnaireResponse,
        CarePlan
        }
    }
% \lstset{language=json,keywordstyle={\bfseries \color{blue}}}
% \lstset{keywordstyle={\bfseries \color{blue}}}
\def\code#1{\texttt{#1}} %format for \code text

\newcommand{\fhir}{FHIR\textsuperscript{\textregistered}}
\newcommand{\hl}{HL7\textsuperscript{\textregistered}}

\newcommand{\tabpatient}{\input{self-reported-nicotine-using-patient.tex}}
\newcommand{\tabcondition}{\input{self-reported-nicotine-usage.tex}}
\newcommand{\tabplan}{\input{nicotine-usage-treatment-plan.tex}}
\newcommand{\tabquestionnaire}{\input{nicotine-treatment-questionnaire.tex}}
\newcommand{\tabresponse}{\input{nicotine-treatment-questionnaire-response.tex}}
\newcommand{\tabcomposition}{\input{exported-nicotine-usage-treatment-data.tex}}

\begin{document}

\tableofcontents

\chapter{Introduction}
\label{ch:intro}

This document establishes a standard for interoperability for Digital Health Applications that treat self-reported nicotine usage. It is based on the internationally recognized \hl \fhir (Fast Healthcare Interoperability Resources) standard for interoperability 
within healthcare \cite{FHIR}. This document is accompanied by FHIR artefacts which are helpful in implementing the standard. The FHIR project and artefacts can be found at the \href{https://simplifier.net/treat-nicotine-usage-diga}{FHIR project page} \cite{project}.
Henceforth in this text, the \textit{Standard} refers to this document in combination with the FHIR artefacts.

\section{Version History}
\begin{table}[H]
    \begin{tabular}{@{}lll@{}}
    \textbf{As per}                 & \textbf{Version}         & \textbf{Changes} \\ \midrule
    \multicolumn{1}{l|}{March 2021} & \multicolumn{1}{l|}{1.0} & Initial release. \\ \bottomrule
    \end{tabular}
    \label{tab:version}
    \end{table}

\section{Objective}
The purpose of the Standard is to establish a standard way of interoperability for Digital Health Applications (DiGAs) that treat self-reported nicotine usage (F17.2 using the ICD-10 coding system \cite{dimdi}). The Standard must be based on internationally recognized
standards for interoperability within healthcare. The Standard should be as flexible as possible to allow for different applications to fit it to their need. The specific use-cases that the Standard is targetting are:

\begin{itemize}
    \item Interoperability between similar DiGAs. DiGAs which treat self-reported nicotine usage are able to exchange an individual patient's data using the format established in the Standard.
    \item Interoperability between a DiGA and journalling systems such as patient health records (ePA). Electronic patient journals are able to extract information using the Standard on what treatment the patient has received through the DiGA and key insights gained by the DiGA on the patient and the patient's condition.
\end{itemize}

\section{Online presence and further development}
The FHIR resources developed for the Standard can be retrieved as a package at the \href{https://simplifier.net/treat-nicotine-usage-diga}{project page} \cite{project}, 
and the accompanying implementation guide which is a shorter version of this document can be found there too \cite{ig}.

The source code for the Standard: this document as well as all FHIR resources and their implementation guide; are open source and can be 
found on \href{https://github.com/alex-therapeutics/diga-nicotine-usage-fhir}{GitHub} \cite{github}. Further development of the Standard
takes place there and anyone may raise issues or suggest changes to it which can be incorporated in future versions.
The project is maintained by Alex Therapeutics.

\section{Copyright and license}

\begin{itemize}
    \item The Standard, which is this document and the attached FHIR artefacts, is licensed under the Creative Commons Attribution 4.0 International License. 
To view a copy of this license, visit \href{http://creativecommons.org/licenses/by/4.0/}{http://creativecommons.org/licenses/by/4.0/} or send a letter to Creative Commons, PO Box 1866, Mountain View, CA 94042, USA.
Copyright for the Standard is held by Alex Therapeutics AB. 
    \item The source code for the Standard is licensed under the Apache 2.0 license. 
You can read a copy of this license in the \href{https://github.com/alex-therapeutics/diga-nicotine-usage-fhir/blob/main/LICENSE}{GitHub project} \cite{github}.
Copyright for the source code is held by Alex Therapeutics AB together with individual contributors to it.
    \item The base \hl \fhir standard which the Standard is derived from is copyrighted by \hl and subject to their \href{http://hl7.org/fhir/license.html}{license} \cite{fhirlic}.
HL7, FHIR and the FHIR [FLAME DESIGN] are the registered trademarks of Health Level Seven International and their use does not constitute endorsement by HL7.
\end{itemize}

\section{Outline}
Chapter \ref{ch:bg} gives a background on central concepts like DiGA, FHIR, and the speficic type of DiGA the Standard is targetting. The specification of the FHIR resources with comments on how they should be implemented follows in Chapter \ref{ch:spec}.
Lastly, there are examples contained in Appendix \ref{app:ex}.


\chapter{Background}
\label{ch:bg}

This chapter contains descriptions and references for central concepts used in the Standard.

\section{Digital Health Application}
A Digital Health Application (DiGA) is a CE-marked medical device based on digital technologies (f.e, a mobile application). Following the German Digital Healthcare Act, they were introduced as "apps on prescription" for German patients and became reimbursable by statutory health insurances \cite{diga}.

To be listed as a DiGA there are several requirements, and the German Federal Institute for Drugs and Medical Devices (BfArM) provides a guide for manufacturers on how to meet them \cite{guide}.
One of the requirements for DiGA is on interoperability. The guide \cite[p.51]{guide} states that:

\begin{quotation}
    \noindent \textit{DiGA should prospectively communicate with each other and interact with other services and applications on the national e-health infrastructure, so that real added value for healthcare can be achieved.}
\end{quotation}

Furthermore, interoperability should be achieved using available standards. If there is no national or international standard available that apply to the specifics of a DiGA, an effort should be made to create a new standard and include it in the directory for IT standards in the German healthcare system, \textit{vesta} \cite{vesta}.
Such a new standard should be based on existing internationally recognized standards, for example by \textit{the combination and extension of several HL7-FHIR profile definitions.} \cite[p.53-54]{guide}.

\section{Fast Healthcare Interoperability Resources}
Fast Healthcare Interoperability Resources (\fhir) \cite{FHIR} is a standard for health care data exchange, published by \hl. It is considered the "next generation" standard for interoperability within healthcare, internationally.
Among the advantages are a strong focus on implementation with multiple existing libraries, flexibility with the possibility to extend and profile existing base resources, a strong foundation in web standards like XML, JSON, etc., and support for RESTful data exchange.

FHIR is made up from base components called \textit{Resources}. An example resource is the \textit{Patient} resource, which contains data about a single patient. 

The base resources aim to cover all data points within healthcare, both clinical and administrative, and because they cover such a wide range of use-cases, the base resources are generic and flexible by nature.
An important feature is then that the base resources can be modified to suit more specific use cases by using \textit{profiling} and \textit{extensions}. A profile on a base resource means constraining it to behave in a certain way.
For example, you could constrain the Patient resource to make it mandatory to provide a birthdate, which is optional in the base resource. You can also add extensions, which are attributes which do not exist yet on the resource. For example, you could
add the extension "smokingStatus" on a Patient to signify is this patient is currently smoking or not. Further information on \fhir can be obtained via their online standard \cite{FHIR}.

\section{German national profile}
\hl Deutschland \cite{hlde} maintains the national \fhir profile for Germany \cite{debasis}. It contains resources which may be useful for the German national healthcare context and a use-case-specific profile which is to be used in Germany should use applicable resources from it.

\section{ICD-10-GM}

The International Statistical Classification of Diseases and Related Health Problems, German Modification (ICD-10-GM) is the official classification for encoding diagnoses in medical care in Germany \cite{icd}.
The codes can be found online at the DIMDI (Deutsches Institut Für Medizinische Dokumentation und Information) website \cite{dimdi}. The codes F17.x deals with tobacco use, and specifically F17.2, which is referenced in the Standard,
is the coding used for addiction syndrome related to tobacco.

% \section{Treating nicotine usage}

\chapter{Specification}
\label{ch:spec}

This chapter contains descriptions of the FHIR resources developed for use by DiGA treating nicotine addiction (F17.2 \cite{dimdi}). It is important to note that the DiGAs that implement this standard do not and should not
make the diagnosis that the patient is addicted to nicotine: that is done previously by a medical professional who prescribes the DiGA or the condition is recognized by the patient him/herself. 
Thus, data references to the medical condition and to status updates of it are explicitly stated to be \textit{self reported}, which refers to this fact.

When building the resources special consideration has been taken to make them as open and flexible as possible, to allow for DiGAs treating nicotine addiction to suit it to their needs, while still delivering a
well-specified structure for exporting a single patient's data (see \ref{sec:export}).


\section{Dependencies}

The profiles in the Standard depend on the core \fhir \cite{FHIR} profiles, and on the German national FHIR profile \cite{debasis}.

\section{Exporting Data}
\label{sec:export}

The Standard deals with exporting patient data concerning their DiGA nicotine addiction treatment in general,
but in particular with exporting a \textit{single patient's} data as one document. This is a DiGA requirement.

When exporting a single patient's data, implementers \textbf{must} use the \code{Bundle} \fhir resource \cite{bundle}. A bundle \textbf{must} contain an \code{ExportedNicotineUsageTreatmentData} resource (see \ref{sec:comp}), which is a profile on \code{Composition},
which defines metadata about what is contained in the bundle and where to find it. Next, the bundle must contain all resources pointed to in the composition. In this way, it is possible
to export all relevant data for a single patient in one document using the Standard. There is an example of this in Appendix \ref{app:ex}.

\section{Artefacts}
The FHIR profiles, extensions, valuesets and codesystems built for the Standard are listed in Table \ref{tab:all}.

\begin{table}[]
    \begin{tabular}{@{}lll@{}}
    \toprule
    \textbf{Name}                                               & \textbf{Parent}                            & \textbf{Type} \\ \midrule
    \multicolumn{1}{l|}{SelfReportedNicotineUsingPatient}       & \multicolumn{1}{l|}{Patient}               & Profile       \\ \midrule
    \multicolumn{1}{l|}{SelfReportedNicotineUsage}              & \multicolumn{1}{l|}{Condition}             & Profile       \\ \midrule
    \multicolumn{1}{l|}{NicotineUsageTreatmentPlan}             & \multicolumn{1}{l|}{CarePlan}              & Profile       \\ \midrule
    \multicolumn{1}{l|}{NicotineTreatmentQuestionnaire}         & \multicolumn{1}{l|}{Questionnaire}         & Profile       \\ \midrule
    \multicolumn{1}{l|}{NicotineTreatmentQuestionnaireResponse} & \multicolumn{1}{l|}{QuestionnaireResponse} & Profile       \\ \midrule
    \multicolumn{1}{l|}{ExportedNicotineUsageTreatmentData}     & \multicolumn{1}{l|}{Composition}           & Profile       \\ \midrule
    \multicolumn{1}{l|}{EffectiveNicotineIntervention}          & \multicolumn{1}{l|}{-}                     & Extension     \\ \midrule
    \multicolumn{1}{l|}{NicotineTrigger}                        & \multicolumn{1}{l|}{-}                     & Extension     \\ \midrule
    \multicolumn{1}{l|}{SelfReportedSmokingStatus}              & \multicolumn{1}{l|}{-}                     & Extension     \\ \midrule
    \multicolumn{1}{l|}{EffectiveInterventionCode}              & \multicolumn{1}{l|}{-}                     & ValueSet      \\ \midrule
    \multicolumn{1}{l|}{SelfReportedSmokingStatusCode}          & \multicolumn{1}{l|}{-}                     & ValueSet      \\ \midrule
    \multicolumn{1}{l|}{TriggerCode}                            & \multicolumn{1}{l|}{-}                     & ValueSet      \\ \midrule
    \multicolumn{1}{l|}{YesOrNo}                                & \multicolumn{1}{l|}{-}                     & ValueSet      \\ \midrule
    \multicolumn{1}{l|}{EffectiveInterventionCodeSystem}        & \multicolumn{1}{l|}{-}                     & CodeSystem    \\ \midrule
    \multicolumn{1}{l|}{SelfReportedStatusCodeSystem}           & \multicolumn{1}{l|}{-}                     & CodeSystem    \\ \midrule
    \multicolumn{1}{l|}{TriggerCodeSystem}                      & \multicolumn{1}{l|}{-}                     & CodeSystem    \\ \bottomrule
    \end{tabular}
    \caption{FHIR resources created for the Standard}
    \label{tab:all}
    \end{table}

The resources can be accessed and downloaded at the \href{https://simplifier.net/treat-nicotine-usage-diga}{project page on Simplifier.net} where there is also an \href{https://simplifier.net/guide/self-reported-nicotine-usage-diga/home}{Implementation Guide} \cite{ig} with brief instructions on how to use the resources.

We will list the artifacts here in the context of how to use them along with figures describing them. Note that we will only show and discuss the differential tables here, the resources otherwise conform to the core FHIR specification and you can view the full resources in snapshot form in the implementation guide.

\subsection{Representing an app user/patient}
\tabpatient

The DiGA user (patient) will be represented using the \code{SelfReportedNicotineUsingPatient} resource (Table \ref{tab:SelfReportedNicotineUsingPatient}). This resource should be used when the DiGA user is not supervised by a medical professional during the DiGA treatment, which means the
patient is \textit{self reporting}.

Since there will be different information available on the patient for different DiGAs, this is a flexible profile allowing all defaults on \code{Patient}. 
There are however two optional extensions to quickly give information on this person's common triggers for nicotine usage and interventions that have been effective for this person.


The extension field \code{commonNicotineTrigger} allows for exporting information about which common triggers there are for a patient's nicotine usage. 
Values for this extension \textbf{should} be from the value set \code{TriggerCode} (fig TODO). If an appropriate code is not available, implementers \textbf{should} use the string value of \code{CodeableConcept}, 
and implementers \textbf{may} raise an issue on GitHub \cite{github} to include their code in the value set in future versions of the Standard.


% TODO figure extension

% TODO figure valueset

% TODO figure codesystem?

The extension field \code{effectiveNicotineIntervention} allows for exporting information about which interventions have been effective for this patient to prevent their use of nicotine.
Values for this extension \textbf{should} be from the value set \code{EffectiveInterventionCode} (fig). If an appropriate code is not available, implementers \textbf{should} use the string value of \code{CodeableConcept}, 
and implementers \textbf{may} raise an issue on GitHub \cite{github} to include their code in the value set in future versions of the Standard.

% TODO figure extension

% TODO figure valueset

% TODO figure codesystem?

\subsection{Representing the condition}
\tabcondition

The profile on \code{Condition} called \code{SelfReportedNicotineUsage} (Table \ref{tab:SelfReportedNicotineUsage}) should be used when the condition is not supervised by a medical professional during the DiGA treatment, but self-reported by the patient.
It allows the DiGA to make it clear what is being treated while at the same time making it clear that the DiGA is not making the diagnosis and that the status of the condition is reported by the patient him/herself. 
To achieve this, the profile disallows clinical attributes like evidence, clinicalStatus etc. It adds an extension called \code{currentSelfReportedSmokingStatus} to allow for a readily accessible way to find out what the current status of the condition is reported to be. 
The condition coding points to the german base profile's \cite{debasis} ICD-10-GM code \cite{icd} for nicotine addiction, F17.2.

% TODO figure extension

% TODO figure valueset & codesystem

The extension field \code{currentSelfReportedSmokingStatus} allows for quickly viewing the status on the condition. It should include the latest report on the condition status made by the patient.
Values for this extension's \code{status} attribute \textbf{should} be from the value set \code{SelfReportedSmokingStatusCode} (fig). If an appropriate code is not available, implementers \textbf{should} use the string value of \code{CodeableConcept}, 
and implementers \textbf{may} raise an issue on GitHub \cite{github} to include their code in the value set in future versions of the Standard.

\subsection{Representing the DiGA care program}
\tabplan
For patient journals and doctors there needs to be a quick way of discovering what the DiGA does or did when the patient was using it. 
Implementers should use the profile on \code{Condition} called \code{NicotineUsageTreatmentPlan} (Table \ref{tab:NicotineUsageTreatmentPlan} for this. 
Furthermore, implementers \textbf{must} use the \code{description} field to write a summary of what the DiGA does as part of its program.
Consumers of the resource should use these fields to gain a brief understanding of what the DiGA does as part of its treatment.
There is an optional extension available, \code{selfReportedSmokingStatus}, to allow for a timeline of how the nicotine usage status of the patient has changed over time, with one record each time the patient has reported a change in the status. 
If the implementer has this data, it should be in included in the data using this extension. The extension used for this is the one defined in Figure \ref{TODO}.

\subsection{Representing app activities}
\tabquestionnaire
\tabresponse
App activities can vary between DiGAs so implementers are free to represent these in various ways including \code{CarePlan}'s \code{activity} field. 
One common demoninator for apps treating nicotine usage however is that the patients are encouraged to track, \textit{log}, their cigarettes. 
Often, the patient also answers questionnaires about various things, to gather information about their nicotine usage and to allow the patient to gain insight themselves on how their addiction works.

To represent these activities, logging and questionnaires, implementers should use \code{NicotineTreatmentQuestionnaire} (Table \ref{tab:NicotineTreatmentQuestionnaire}) to define how the questionnaire is presented in the application, 
and \code{NicotineTreatmentQuestionnaireResponse} (Table \ref{tab:NicotineTreatmentQuestionnaireResponse}) to represent an individual patient's response to it. It is required to give a \code{titte}, \code{description} and \code{purpose} to the questionnaire to explain what it does.
Implementers are encouraged to use the \code{TriggerCode} and \code{EffectiveInterventionCode} value sets when they are appropriate as answers (TODO ref to tables).

\subsection{Mandatory metadata for exports}
\label{sec:comp}
\tabcomposition
When exporting data for a single patient in a \code{Bundle}, as described in \ref{sec:export}, there must be metadata present. 
For this, a profile on \code{Composition} is used called \code{ExportedNicotineUsageTreatmentData} (Table \ref{tab:ExportedNicotineUsageTreatmentData}).
The \code{section} field is sliced to contain sections for all resources required to export a single patient's data: 
\code{patientData}, \code{selfReportedCondition}, \code{nicotineUsageTreatmentPlan}, \code{questionnaires}, and \code{questionnaireResponses}.
Each of these must contain a reference to a resource contained in the \code{Bundle}. For example, the section slice \code{selfReportedCondition} must
contain a reference to the patient's \code{SelfReportedNicotineUsage} condition.

\printbibliography[heading=bibintoc, title={References}]


\appendix

\chapter{Examples}
\label{app:ex}

For completeness, we include one full example of a \code{Bundle} export containing one patient's complete set of data exported according to the Standard.
This example along with individual examples of each profile is available online at the \href{https://simplifier.net/treat-nicotine-usage-diga}{project page on Simplifier.net} \cite{project}.

\lstinputlisting[breaklines]{../fhir-profile/output/Bundle-PatientExportBundleExample.xml}
% \lstinputlisting{../fhir-profile/output/Bundle-PatientExportBundleExample.json}


\end{document}